---
title: "Deploying your app"
author: "Roy Francis"
date: "`r format(Sys.time(), '%d-%b-%Y')`"
output:
  html_document:
    theme: flatly
    highlight: tango
    number_sections: true
template:
  bootstrap:5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

In this vignette, we look at various ways to deploy your app to the end user.

## R

The app directory with all the files can be archived (.zip, .tar.gz etc.) and shared. The receiver will need to install R as well as all the dependency packages (listed on home page) and then launch the app by running `shiny::runApp()` in the app directory.

## Docker

A better solution is to package the shiny app into a docker image. In this example, we will create an app and containerize it into a docker image.

```{r,eval=TRUE,message=FALSE,warning=FALSE}
library(easyshiny)
library(Seurat)
library(SeuratData)

ifnb <- LoadData("ifnb")
ifnb <- ifnb |>
  NormalizeData() |>
  ScaleData() |>
  FindVariableFeatures() |>
  RunPCA() |>
  RunUMAP(dims=1:20)

ifnb
```

```
An object of class Seurat 
14053 features across 13999 samples within 1 assay 
Active assay: RNA (14053 features, 2000 variable features)
 2 dimensional reductions calculated: pca, umap
```

After pre-processing (normalisation, scaling, PCA, UMAP), we can proceed to building the app.

```{r}
ifnb_conf <- create_config(ifnb)
make_app(ifnb, ifnb_conf, gex.assay="RNA", gex.slot="data", shiny.title = "IFNB", shiny.dir="ifnb")
shiny::runApp("ifnb")
```

Once you have verified that the app works and made any changes to the app as needed, you can create a Dockerfile using the function `make_dockerfile()`.

```{r}
make_dockerfile(path="ifnb")
```

This will write out a Dockerfile into the same directory as the app.

<div style="color: #8a6d3b;background-color: #fcf8e3;border-color:#faebcc; padding:6px; border-radius:4px; margin: 1em 0;" style="block">
<p style="color: #8a6d3b;background-color: #fcf8e3;">
&nbsp; <i class="fas fa-exclamation-circle"></i> &nbsp; If using GitHub packages, images are pushed to the personal or org account and not to a specific repository. To link packages to a repository, add the following to the Dockerfile:
</p>

```
LABEL org.opencontainers.image.source https://github.com/OWNER/REPO
```
</div>

In the terminal, change to the app directory and build the image using the syntax `docker build -t <image-name> .`. My docker image will be named `shiny-ifnb`. You can name it [anything](https://docs.docker.com/engine/reference/commandline/tag/) you like. Note that everything in the app folder will be copied into the image.

```{bash}
cd ifnb
docker build -t shiny-ifnb .
```

This will take a while to build and when done, you can list all images and find your app:

```{bash}
docker images
```

Launch the app and it should be available in the browser at `http://0.0.0.0:3838`.

```{bash}
docker run --rm -p 3838:3838 shiny-ifnb
```

### Distributing the image

The docker image can be distributed through image repositories. The docker image can be run through the terminal and also graphically through [Docker Desktop](https://www.docker.com/products/docker-desktop/) on Linux, Mac and Windows.

Two options for distributing the docker image is discussed: [**DockerHub**](https://hub.docker.com/) and [**GitHub packages**](https://github.com/features/packages).

#### DockerHub

<p style="color: #8a6d3b;background-color: #fcf8e3;border-color:#faebcc;padding: 6px;border-radius: 4px;" style="inline-block">
&nbsp; <i class="fas fa-exclamation-circle"></i> &nbsp; Be mindful of access permissions to the image, since the app/image contains the expression data and metadata.
</p>

Tag the image as needed with repo, version etc using the syntax `docker tag <image-name> <hub-name>/<repo-name>:<tag>`.

```{bash}
docker tag shiny-ifnb royfrancis/shiny-ifnb:v1.0
docker tag shiny-ifnb royfrancis/shiny-ifnb:latest
```

To push to DockerHub, use the following syntax: `docker push <hub-name>/<repo-name>:<tag>`

```{bash}
docker login docker.io
docker push royfrancis/shiny-ifnb:v1.0
docker push royfrancis/shiny-ifnb:latest
```

Once upload is complete, you can choose to remove the local image:

```{bash}
docker rmi shiny-ifnb royfrancis/shiny-ifnb:v1.0 royfrancis/shiny-ifnb:latest
docker images
```

Then try to run the app:

```{bash}
docker run --rm -p 3838:3838 royfrancis/shiny-ifnb:latest
```

This should fetch the image from dockerhub and launch the app. The should be available at `http://0.0.0.0:3838`.

#### GitHub packages

<p style="color: #8a6d3b;background-color: #fcf8e3;border-color:#faebcc;padding: 6px;border-radius: 4px;" style="inline-block">
&nbsp; <i class="fas fa-exclamation-circle"></i> &nbsp; Note that SciLifeLab Serve does not support GitHub packages yet.
</p>

GitHub packages are pushed to a personal or organisation account and not to a specific repository. To link packages to a repository, add the following to the Dockerfile:

`LABEL org.opencontainers.image.source https://github.com/OWNER/REPO`

Create a personal access token [here](https://github.com/settings/tokens). Select **Tokens (Classic)** and check permissions **read:packages** and **write:packages**.

Login to github package repository.

```{bash}
echo "personalaccesstoken" | docker login ghcr.io -u githubusername --password-stdin
```

To push to GitHub packages, tag the image using the following syntax: `docker push <hub-name>/<user|org>/<image-name>:<tag>`

```{bash}
docker tag royfrancis/shiny-ifnb ghcr.io/nbisweden/shiny-ifnb:latest
docker push ghcr.io/nbisweden/shiny-ifnb:latest
```

To launch the app:

```{bash}
docker run --rm -p 3838:3838 ghcr.io/nbisweden/shiny-ifnb:latest
```

## Singularity

The docker image can be easily converted to a singularity image.

```{bash}
docker save ghcr.io/nbisweden/shiny-ifnb:latest -o shiny-ifnb.tar
singularity build --sandbox shiny-ifnb.tmp docker-archive://shiny-ifnb.tar
singularity build shiny-ifnb.sif shiny-ifnb.tmp

singularity run shiny-ifnb.sif
```

## Hosting

### ShinyApps.io

The shiny app can be hosted on shinyapps.io directly from the app files using R package [rsconnect](https://shiny.rstudio.com/articles/shinyapps.html).

### SciLifeLab Serve

<p style="color: #8a6d3b;background-color: #fcf8e3;border-color:#faebcc;padding: 6px;border-radius: 4px;" style="inline-block">
&nbsp; <i class="fas fa-exclamation-circle"></i> &nbsp; Note that SciLifeLab Serve does not support GitHub Packages yet. The image must be on DockerHub.
</p>

SciLifeLab users have the possibility of hosting the docker image on [SciLifeLab Serve](https://serve.scilifelab.se/). Login and create a new **Project**. Then **Serve > New > Shiny App**. Fill in the form:

```
Name: IFNB scRNA-Seq
Description: IFNB single-cell transcriptomics data
Subdomain: ifnb.serve.scilifelab.se
Permissions: Project
Persistent Volume: project-vol
App port: 3838
Image: royfrancis/shiny-ifnb
```

For subdomain, **ifnb** was added as a new domain to have a persistent URL. The `Image` refers to DockerHub username and repo. `Permissions` sets app access to those registered to the **Project** or completely open to **Public**.

The app will be available at [https://ifnb.serve.scilifelab.se/](https://ifnb.serve.scilifelab.se/).

### Cloud

Docker images are versatile and can be hosted through most cloud providers such as [Amazon ECS](https://aws.amazon.com/ecs/), [Google Cloud Run](https://cloud.google.com/run), [Microsoft Azure](https://azure.microsoft.com/en-gb/products/kubernetes-service/docker/), [Vultr](https://www.vultr.com/apps/docker) etc.
