---
title: "Deploying your app"
author: "Roy Francis"
date: "`r format(Sys.time(), '%d-%b-%Y')`"
output:
  html_document:
    theme: flatly
    highlight: tango
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

In this vignette, we look at various ways to deploy your app to the end user.

## R

The app directory with all the files can be archived (ie; .zip, .tar.gz etc.) and shared. The receiver will need to install R as well as all the dependency packages (listed on home page) to be able to launch the app by running `shiny::runApp()` in the app directory.

## Docker

A better solution is to package the shiny app into a docker image. In this example, we will create an app and containerize it into a docker image.

```{r,eval=TRUE,message=FALSE,warning=FALSE}
library(easyshiny)
library(Seurat)
library(SeuratData)

ifnb <- LoadData("ifnb")
ifnb <- ifnb |>
  NormalizeData() |>
  ScaleData() |>
  FindVariableFeatures() |>
  RunPCA() |>
  RunUMAP(dims=1:20)

ifnb
```

```
An object of class Seurat 
14053 features across 13999 samples within 1 assay 
Active assay: RNA (14053 features, 2000 variable features)
 2 dimensional reductions calculated: pca, umap
```

After pre-processing (normalisation, scaling, PCA, UMAP), we can proceed to building the app.

```{r}
ifnb_conf <- create_config(ifnb)
make_app(ifnb, ifnb_conf, gex.assay="RNA", gex.slot="data", shiny.title = "IFNB", shiny.dir="ifnb")
shiny::runApp("ifnb")
```

Once you have verified that the app works and made any changes to the app as needed, you can create a Dockerfile using the function `make_dockerfile()`.

```{r}
make_dockerfile(path="ifnb")
```

This will write out a Dockerfile into the same directory as the app.

**Note!** If using GitHub packages, images are pushed to the personal or org account and not to a specific repository. To link packages to a repository, add the following to the Dockerfile:

`LABEL org.opencontainers.image.source https://github.com/OWNER/REPO`

In the terminal, change to the app directory and build the image using the syntax `docker build -t <image-name> .`. Note that everything in the app folder will be copied into the image.

```{bash}
cd ifnb
docker build -t shiny-ifnb .
```

This will take a while to build and when done, you can list all images and find your app:

```{bash}
docker images
```

Launch the app and it should be available in the browser at `http://0.0.0.0:8787`.

```{bash}
docker run --rm -p 8787:8787 shiny-ifnb
```

### Distributing the image

The docker image can be distributed through image repositories. The docker image can be run through the terminal and also graphically through Docker Desktop on Mac and Windows.

I will discuss two options: DockerHub and GitHub packages.

#### DockerHub

**Note!** Be mindful of sensitive data since the app/image contains the expression data and metadata.

Tag the image as needed with repo, version etc using the syntax `docker tag <image-name> <hub-name>/<repo-name>:<tag>`.

```{bash}
docker tag shiny-ifnb royfrancis/shiny-ifnb:v1.0
docker tag shiny-ifnb royfrancis/shiny-ifnb:latest
```

To push to DockerHub, use the following syntax: `docker push <hub-name>/<repo-name>:<tag>`

```{bash}
docker login docker.io
docker push royfrancis/shiny-ifnb:v1.0
docker push royfrancis/shiny-ifnb:latest
```

Once upload is complete, you can choose to remove the local image:

```{bash}
docker rmi shiny-ifnb royfrancis/shiny-ifnb:v1.0 royfrancis/shiny-ifnb:latest
docker images
```

Then try to run the app:

```{bash}
docker run --rm -p 8787:8787 royfrancis/shiny-ifnb:latest
```

This should fetch the image from dockerhub and launch the app. The should be available at `http://0.0.0.0:8787`.

#### GitHub packages

GitHub packages are pushed to the personal or org account and not to a specific repository. To link packages to a repository, add the following to the Dockerfile:

`LABEL org.opencontainers.image.source https://github.com/OWNER/REPO`

Create a personal access token [here](https://github.com/settings/tokens). Select **Tokens (Classic)** and check **read:packages** and **write:packages**.

Login to github package repository.

```{bash}
echo "personalaccesstoken" | docker login ghcr.io -u githubusername --password-stdin
```

To push to GitHub packages, tag the image using the following syntax: `docker push <hub-name>/<user|org>/<image-name>:<tag>`

```{bash}
docker tag royfrancis/shiny-ifnb ghcr.io/nbisweden/shiny-ifnb:latest
docker push ghcr.io/nbisweden/shiny-ifnb:latest
```

To launch the app:

```{bash}
docker run --rm -p 8787:8787 ghcr.io/nbisweden/shiny-ifnb:latest
```

## Singularity

The docker image can be easily converted to a singularity image.

```{bash}
docker save ghcr.io/nbisweden/shiny-ifnb:latest -o shiny-ifnb.tar
singularity build --sandbox shiny-ifnb.tmp docker-archive://shiny-ifnb.tar
singularity build shiny-ifnb.sif shiny-ifnb.tmp

singularity run shiny-ifnb.sif
```

## Hosting

### ShinyApps.io

The shiny app can be hosted on shinyapps.io directly from the app files using R package [rsconnect](https://shiny.rstudio.com/articles/shinyapps.html).

### SciLifeLab Serve

SciLifeLab users have the possibility of hosting the docker image on [SciLifeLab Serve](https://serve.scilifelab.se/). Login and create a new **Project**. Then **Serve > New > Shiny App**. Fill in the form:

```
Name: IFNB scRNA-Seq
Description: IFNB single-cell transcriptomics data
Subdomain: ifnb.serve.scilifelab.se (Add ifnb as a new domain)
Permissions: Project
Persistent Volume: project-vol
App port: 8787
Image: ghcr.io/nbisweden/shiny-ifnb:latest
```

### Cloud

Docker images are versatile and can be hosted through most cloud providers such as [Amazon ECS](https://aws.amazon.com/ecs/), [Google Cloud Run](https://cloud.google.com/run), [Microsoft Azure](https://azure.microsoft.com/en-gb/products/kubernetes-service/docker/), [Vultr](https://www.vultr.com/apps/docker) etc.